Modulations of gravitational-wave signals: a Fourier-domain approach for the response of LISA-type detectors and for precessing binaries

### 2017-01-26 docs/paper_method.tex
Update Makefile for Datura, and small changes to the paper.

With the recent detections of gravitational-wave signals emitted by the coalescences of binary black holes, announced by the LSC collaboration after the first obervation run of the two LIGO instruments~\cite{LIGO-theevent-2016,LIGO-christmasevent-2016,LIGO-O1BBH-2016}, gravitational-wave astronomy has entered its observational era. With the recent start of LIGO's second observation run, and later the extension of the ground-based detectors network with Virgo~\cite{Virgo}, KAGRA~\cite{KAGRA} and LIGO-India~\cite{INDIGO}, many more detections from compact object coalescences are to come, with binary black holes being expected to be the dominant sources. Moreover, the European Space Agency has issued a mission call for its L3 mission, adressing the science theme ``The Gravitational Universe'' with a tentative launch in 2034, possibly opening the way for a space-borne observatory based on the concept of the Laser Interferometer Space Antenna (LISA)~\cite{LISA09}. Among other sources of gravitational waves, LISA will be able to detect and characterize binary black hole coalescences to cosmological distances over a wide range of masses, from high-redshift observations of supermassive black hole binaries with $M\sim 10^{7} \Msol$ down to the observation of LIGO-type sources with $M\sim 10^{2} \Msol$~\cite{Sesana16}.

The Bayesian analysis and parameter recovery of gravitational-wave signals may require millions of evaluations of waveform templates to sample the posterior probability distribution, which poses a computational challenge for templates including consistently the inspiralling phase of the evolution, the merger and the ringdown (IMR). State-of-the-art IMR templates incorporate information from both the perturbative results of post-Newtonian theory covering the inspiral (see e.g.~\cite{BlanchetLiving}) and from numerical relativity simulations covering the end of the inspiral and the merger-ringdown phase~\cite{}. The two main approaches currently used to produce complete waveform models are (i) the Effective-One-Body (EOB) approach~\cite{BD99}, which incorporates a resummation of post-Newtonian results and a calibration to numerical relativity, and which has been extended to include both spins of the two bodies~\cite{Taracchini+13, Pan+13, Bohe+16} (ii) phenomenological templates, which extend post-Newtonian signals by postulating a Fourier-domain analytic expression for the amplitude and phase and by fitting coefficients against numerical relativity waveforms~\cite{Husa+15, Khan+15}, and which include a simplified model for precession~\cite{Hannam+13}. The second type of models produce directly the amplitude and the phase of the Fourier-domain waveform in the form of a piecewise analytic fit. EOB waveforms, by contrast, require to integrate ordinary differential equations in the time-domain, which can be time-consuming. For the non-precessing case, Reduced Order Models (ROM), also called surrogate models, have been developed to speed up considerably the waveform generation, directly in the Fourier domain, without losing accuracy~\cite{Field+13, Puerrer14, Bohe+16}.

A complete representation of the spin effects in fast IMR templates still remains a frontier of gravitational wave signal modelling. In presence of misaligned spins, the system will endure precession of the orbital plane as it evolves, leading to a modulation of the signal that is more pronunced for observers that see the system as close to edge-on~\cite{ACST94}. Precessing EOB templates can represent all six degrees of freedom of the two spins~\cite{Pan+13, BTB16} but still require a costly time-domain integration; while precessing phenomenological templates~\cite{Hannam+13} use effective spin parameters representing only the dominant effects of the spins. To model the effect of precession on the emitted waveform, it has been proposed~\cite{BCV03b, BCPTV05, Schmidt+10, OShaughnessy+11, Boyle+11} to decompose precessing waveforms by performing a time-dependent rotation, following the precession of the orbital plane. The resulting waveform in the rotated frame is then close to a non-precessing waveform, an approximation that is used both to build the inspiral part of precessing EOB waveforms~\cite{Pan+13} and in the construction of precessing phenomenological waveforms~\cite{Hannam+13}. To carry over this approach to Fourier-domain waveforms, one needs to understand how to translate the time-domain modulation created by the frame rotation into a Fourier-domain modulation. The most straightforward method consists in using a time-to-frequency correspondence such as the one given in the Stationary Phase Approximation (SPA), a good approximation for the inspiralling phase. This is what is used implicitly in Ref.~\cite{Hannam+13}, with the difference that directly frequency-dependent expressions for the modulation are used, which are then extended to also cover high frequencies in the merger part of the signal. In Ref.~\cite{Klein+14}, a formalism is introduced to go beyond the SPA and compute more accurately the modulation in the Fourier domain; however, this formalism still relies on the SPA for the underlying signal. How to implement

Another frontier of gravitational wave signal modelling is prospective parameter recovery for LISA observations. In the past, LISA data analysis studies for comparable-mass binaries used mainly simplified, inspiral-only models~\cite{}, and Fisher-matrix estimates instead of a full Bayesian analysis. In Ref.~\cite{Klein+15}, an improved model was used, that included the effects of both spins including the corrections beyond the SPA investigated in~\cite{Klein+14}; the limitation to inspiral-only waveforms was compensated by SNR reweighting using full IMR waveforms. The response of the LISA instrument, by contrast with the response of ground-based detectors, has the specificity of being time-dependent. For a signal accumulating over several months, the motion and change of orientation of the detector constellation along its orbit lead to time variability in the response of the detector, which is crucial for the localization of the gravitational-wave source in the sky. More precisely, these effects can be separated as a modulation coming from the rotation of the constellation, and a delay coming from the motion of the spacecrafts in the wavefront of the incoming wave.

@@

\subsection{Waveforms in the Fourier domain and stationary phase approximation}
\label{subsec:SPA}

@@

We recall below an approximation widely used to model gravitational waves signals emitted by inspiralling binaries, the stationary phase approximation (thereafter SPA). It applies in general to chirping signals, and we refer the reader to~\ref{} for details. One first writes the time-domain signal in amplitude and phase form as $h(t) = a(t) e^{-2i\varphi(t)}$, where for gravitational wave signals $\varphi$ will correspond to the orbital phase, with the orbital frequency being $\omega = \dot{\varphi}$. In order to keep close to the notations used in the gravitational wave literature, we introduced a factor of 2 in the phase of the wave, which is appropriate for the dominant 22 harmonic of the signal. The approximation then applies to signals verifying the conditions
%
\be\label{eq:conditionSPA}
	\left| \frac{\dot{a}/a}{\omega} \right| \ll 1\,, \quad \left|\frac{\dot{\omega}}{\omega^{2}} \right| \ll 1\,, \quad \left| \frac{(\dot{a}/a)^{2}}{\dot{\omega}} \right| \ll 1 \,.
\ee
%
Since the integral~\eqref{eq:defFT} defining the Fourier transform is rapidly oscillatory unless the term $2\pi f t$ cancels the evolution of $-2\varphi(t)$, its support is well centered around the point of stationary phase. This determines the time-to-frequency correspondence in the stationary phase approximation, and leads to the definition of this time as an implicit function of frequency by the relation
%
\be\label{eq:deftfSPA}
	\omega(t_{f}^{\rm SPA}) = \pi  f \,.
\ee
% 
For a chirping signal of increasing phase, $\omega>0$ and $\dot{\omega}>0$, there is a unique point of stationary phase, located in the positive frequency range $f>0$. Using the conditions above, one can formally expand the signal around $t_{f}^{\rm SPA}$ to quadratic order in time, according to
\be
	\tilde{h}_{\rm SPA} (f) \simeq a(t_{f}^{\rm SPA}) \exp\left[2i\pi f t_{f}^{\rm SPA}-2i\varphi(t_{f}^{\rm SPA}) \right]  \int \ud t \, e^{-i \dot{\omega} (t_{f}^{\rm SPA}) (t-t_{f}^{\rm SPA})^{2}} \,.
\ee
Note that to be able to treat the amplitude as a constant in the integral above, we used the third condition in~\eqref{eq:conditionSPA}. The resulting complex Gaussian integral yields\footnote{Note that the expressions given here are valid for the $22$ mode of the waveform.}
\begin{subequations}
\begin{align}
	\tilde{h}_{\rm SPA}(f) &= A_{\rm SPA}(f) e^{-i\Psi_{\rm SPA}(f)} \,, \\
	A_{\rm SPA}(f) &= a(t_{f}^{\rm SPA}) \sqrt{\frac{\pi}{\dot{\omega}(t_{f}^{\rm SPA})}} \,, \\
	\Psi_{\rm SPA}(f) &= 2\varphi(t_{f}^{\rm SPA}) - 2\pi f t_{f}^{\rm SPA} + \frac{\pi}{4} \,.
\end{align}
\end{subequations}
Ref.~\cite{Droz+99} evaluated the first correction to this approximation, within the context of post-Newtonian signals, and found that it can be considered as a term of the fifth post-Newtonian order, beyond the accuracy level of our current best models~\cite{BlanchetLiving}.

To understand the separation of timescales in the problem, it will be useful to have at hand the leading-order scaling laws for an inspiral (at the Newtonian order in the PN language). Although inaccurate for the purpose of waveform modelling, these Newtonian estimates will give a good qualitative description of the order of magnitude of the relevant timescales in the inspiral. For a binary with masses $m_{1}, m_{2}$, we define the total mass $m=m_{1}+m_{2}$ and the symmetric mass ratio $\nu = m_{1}m_{2}/m^{2}$. Introducing a time of coalescence $t_{c}$, the relations between the orbital frequency and phase and the time to coalescence $t_{c} - t$ are then given by
\begin{subequations}
\begin{align}
	\omega(t) &= \left[ \frac{256\nu}{5c^{5}} (Gm)^{5/3} (t_{c}-t) \right]^{-3/8} \,, \\
	\varphi(t) &= -\left[ \frac{c^{3}}{G m \nu^{3/5}} (t_{c}-t) \right]^{5/8} \,.
\end{align}
\end{subequations}
For this Newtonian inspiral, applying the SPA gives for the $22$ mode $\tilde{h}(f) = A_{\rm N}(f)e^{-i\Psi_{\rm N}(f)}$ with
%
\begin{subequations}
\begin{align}
	A_{N}(f) &= \frac{G^{2}m^{2} \pi}{Dc^{5}} \sqrt{\frac{2\nu}{3}} v^{-7/2}\,,\\
	\Psi_{\rm SPA}^{\rm N}(f) &= 2\pi f t_{0} - \phi_{0} + \frac{3}{128\nu v^{5}} \,, 
\end{align}
\end{subequations}
%
where $v=(G m \pi f/c^{3})^{1/3}$, $D$ is the luminosity distance to the observer, and where $t_{0}, \phi_{0}$ are constants. It is also customary to rewrite the above relations in terms of the chirp mass $\calM \equiv m\nu^{3/5}$, which is the only mass combination arising in the phase at the leading PN order.

@@

Defining
%
\be
	h_{d}(t) = h(t+d(t)) \,, \quad s(t) = F(t)h_{d}(t) \,,
\ee
%
we have
%
\be
	h_{d}(t) = \int \ud f \, e^{-2i\pi f (t+d(t))}\tilde{h}(f) \,,
\ee
%
and
%
\begin{align}
	\tilde{s}(f) &= \int \ud t \, e^{2i\pi f t} F(t)  \int \ud f' \, e^{-2i\pi f' (t+d(t))}\tilde{h}(f') \nn\\
	&= \int \ud f' \, \tilde{h}(f-f') \int \ud t \, e^{2i\pi f' t} e^{-2i\pi (f-f') d(t)} F(t) \,.
\end{align}
%
We can rewrite the last equation as a kind of convolution with a frequency-dependent Kernel, according to
%
\be\label{eq:FDkernel}
	\tilde{s}(f) = \int \ud f' \, \tilde{h}(f-f') \tilde{g}(f-f',f') \,,
\ee
%
where we introduced the frequency-dependent function of time $g(f,t)$, and its Fourier transform in the auxiliary frequency $f'$, denoted by $\tilde{g}(f,f')$, as
%
\begin{subequations}\label{eq:defg}
\begin{align}
	g(f,t) &= e^{-2i\pi f d(t)} F(t) \,, \\
	\tilde{g}(f,f') &= \int \ud t \, e^{2i\pi f' t} e^{-2i\pi f d(t)} F(t) \,.
\end{align}
\end{subequations}
%
In the case of precessing binaries, where the delays $d(t)$ will be absent, the frequency-dependence of $g$ will vanish and this function will simply reduce to the modulation $F(t)$.

In our approach, two assumptions will be crucial to the approximation:
\begin{itemize}
	\item the time scales entering the signal $h(t)$ are well separated from the ones entering the modulation and the delay, so that $\tilde{g}(f-f',f')$ has support limited to $[-f_{\rm max},f_{\rm max}]$, with $f_{\rm max}$ small compared the characteristic scale for the variation of the Fourier-domain signal;
	\item the original signal has a smooth amplitude and phase in the Fourier-domain, slowly varying over the scale $f_{\rm max}$, so that a local treatment in $f'$ of the convolution in~\eqref{eq:FDkernel} will be possible.
\end{itemize}
Although well verified for the case of LISA-type detectors, the first assumption can be a limitation of the approximation in the case of preccesing binaries, and we will carefully explore its validity in Sec.~\ref{}.

@@

Plugging this relation in~\eqref{eq:FDkernel}, we obtain
\begin{align}
	\tilde{s}(f) &= \tilde{h}(f) \int \ud f' \, \exp\left[ i f' \frac{\ud \Psi}{\ud f} \right] \tilde{g}(f,f') \nn\\
	&= \tilde{h}(f) g\left( f, -\frac{1}{2\pi} \frac{\ud \Psi}{\ud f} \right) \,,
\end{align}
which motivates the introduction of an effective time, seen as a function of frequency,
\be\label{eq:deftf}
	\tf \equiv -\frac{1}{2\pi} \frac{\ud \Psi}{\ud f} \,,
\ee
This definition is a generalization of the time-to-frequency correspondence at the heart of the SPA, as will be discussed in Sec.~\ref{subsec:linkSPA}, but it is worth noting that here it is defined entirely from the Fourier-domain waveform. Fig.~\ref{fig:tf} shows the time function $\tf$ for the example of an equal-mass waveform. In particular, one should note that $\tf$ is not monotonously increasing with frequency anymore when reaching in the high-frequency part of the waveform, corresponding to the ringdown. This would forbid an unambiguous definition of an inverse fequency-of-time correspondence $f(t)$.

@@

Using this notation, we have simply
%
\be\label{eq:local}
	\tilde{s}_{\rm local}(f) = \tilde{h}(f) g(f, \tf) \,.
\ee
%
The interpretation of this approximation is straightforward: the signal is simply multiplied by the response function evaluated at the time $\tf$. Additionally, when delays are present, the frequency-dependence of the response is evaluated directly at $f$. For precessing binary, this corresponds to the approximation used for the PhenomP approximant~\cite{Hannam+13}, as will be discussed in Sec.~\ref{}.

It is worth noting that a shift in time of the time-domain signal will, by construction, be appropriately propagated to~\eqref{eq:local}. If $h'(t) = h(t -  \Delta t)$ [but the modulation and delays are left unchanged (which is relevant for space-based detectors, not for precessing binaries)], $\Psi'(f) = \Psi(f) - 2\pi f \Delta t$, so that
%
\be\label{eq:localshifted}
	\tilde{s}'_{\rm local}(f) = \tilde{h}'(f) g(f, \tf + \Delta t) \,. 
\ee
%
This observation also justifies that the first derivative of the Fourier-domain phase cannot, in general, be considered a small quantity, as it can take arbitrary values for arbitrary shifts in time of the signal. The leading order approximation corresponds to neglecting the curvature in $\Psi(f)$, and yields a result~\eqref{eq:local} where the passage to the Fourier domain is done by simply using a time-to-frequency correspondence~\eqref{eq:deftf}, which motivates the name ``local approximation''.

@@

We will formally Taylor-expand $\tilde{h}(f-f')$ in $f'$ as follows: we write
%
\begin{subequations}
\begin{align}
	\Psi(f-f') &= \Psi(f) + 2\pi f' \tf + \sum\limits_{p\geq 2}^{N} \frac{(-1)^{p}}{p!} {f'}^{p} \frac{\ud^{p} \Psi}{\ud f^{p}} \,, \label{eq:expandPsi}\\
	A(f-f') &= A(f) \sum\limits_{q\geq 0} \frac{(-1)^{q}}{q!} {f'}^{q} \frac{1}{A}\frac{\ud^{q} A}{\ud f^{q}} \,, \label{eq:expandA}\\
	\tilde{g}(f-f', f') &= \sum\limits_{r\geq 0} \frac{(-1)^{r}}{r!} {f'}^{r} \frac{\partial^{r} }{\partial f^{r}}  \tilde{g}(f,f') \label{eq:expandg}
\end{align}
\end{subequations}
%
This expansion will only be usable in practice when we will be able to keep only a few terms in each of these expansions. We therefore limited the expansion in $\Psi$ to a finite order $N$, to keep the formula readable. Notice that we expand $\tilde{g}(f-f',f')$ in $f'$ only in its first argument. We can also interchange the $f$-derivatives of $g$ with its Fourier transform.

>>>>

With the recent detections of gravitational-wave signals emitted by the coalescences of binary black holes, announced by the LSC collaboration after the first obervation run of the two LIGO instruments~\cite{LIGO-theevent-2016,LIGO-christmasevent-2016,LIGO-O1BBH-2016}, gravitational-wave astronomy has entered its observational era. With the recent start of LIGO's second observation run, and later the extension of the ground-based detectors network with Virgo~\cite{Virgo}, KAGRA~\cite{KAGRA} and LIGO-India~\cite{INDIGO}, many more detections from compact object coalescences are to come, with binary black holes being expected to be the dominant sources~\cite{}. Moreover, the European Space Agency has issued a mission call for its L3 mission, adressing the science theme ``The Gravitational Universe'' with a tentative launch around 2034, possibly opening the way for a space-borne observatory based on the concept of the Laser Interferometer Space Antenna (LISA)~\cite{LISA09}. Among other sources of gravitational waves, LISA will be able to detect and characterize binary black hole coalescences to cosmological distances over a wide range of masses, from high-redshift observations of supermassive black hole binaries with $M\sim 10^{7} \Msol$ down to the observation of LIGO-type sources with $M\sim 10^{2} \Msol$~\cite{Sesana16}.

Data analysis for gravitational-wave observations of compact binary coalescences require accurate models (or templates) for the signals, both for ensuring efficient detections of signals that may be buried in instrumental noise, and to extract the physical parameters of the source in subsequent analysis. This has prompted an important and ongoing effort to improve gravitational wave templates. In parallel of the improvement of the accuracy of the templates, an crucial element of the analysis is the computational efficiency of template generation, as Bayesian analysis for parameter estimation of gravitational-wave signals may require millions of evaluations of the likelihood function to sample the posterior probability distribution. Two important axis of improvement of the templates are the coverage of all phases of the evolution of the system, with the inspiral, merger and ringdown (IMR), and the inclusion of the effects of the spins of the compact objects.

State-of-the-art IMR templates need to combine information from the perturbative results of post-Newtonian theory covering the inspiral (see e.g.~\cite{BlanchetLiving}) and from numerical relativity simulations covering the end of the inspiral and the merger-ringdown phase (see e.g.~\cite{Pfeiffer12}). The Effective-One-Body (EOB) approach~\cite{BD99} does so by incorporating a resummation of post-Newtonian results and a calibration to numerical relativity, and has been extended to include both spins of the two bodies~\cite{Taracchini+13, Pan+13, Bohe+16}. Phenomenological templates, which extend post-Newtonian signals by postulating a Fourier-domain analytic expression for the amplitude and phase and by fitting coefficients against numerical relativity waveforms~\cite{Husa+15, Khan+15}, and can include a simplified model for precession~\cite{Hannam+13}. The second type of models produce directly the amplitude and the phase of the Fourier-domain waveform in the form of a piecewise analytic fit. EOB waveforms, by contrast, require to integrate ordinary differential equations in the time-domain, which can be time-consuming. For the non-precessing case, Reduced Order Models (ROM), also called surrogate models, have been developed to speed up considerably the waveform generation, without losing accuracy~\cite{Field+13, Puerrer14, Bohe+16}. Both approaches are therefore able to produce efficiently IMR waveforms. An important aspect is that the resulting waveforms can be represented in the Fourier domain by an amplitude and phase for each mode, with only a few hundred samples~\cite{Puerrer14}.

A complete representation of the spin effects in fast IMR templates still remains a frontier of gravitational wave signal modelling. In presence of misaligned spins, the system will endure a precession of the orbital plane as it evolves, leading to modulations of the signal~\cite{ACST94}. Precessing EOB templates can represent all six degrees of freedom of the two spins~\cite{Pan+13, BTB16} but still require a costly time-domain integration; while precessing phenomenological templates~\cite{Hannam+13} decrease the dimensionnality by using effective spin parameters representing only the dominant effects of the spins. To model the effect of precession on the emitted waveform, it has been proposed~\cite{BCV03b, BCPTV05, Schmidt+10, OShaughnessy+11, Boyle+11} to decompose precessing waveforms by performing a time-dependent rotation, following the precession of the orbital plane. The resulting waveform in the rotated frame is then close to a non-precessing waveform, an approximation that is used in the construction of the inspiral part of precessing EOB waveforms~\cite{Pan+13} and in the construction of precessing phenomenological waveforms~\cite{Hannam+13}. For Fourier-domain waveforms, one needs to understand how to translate the time-domain modulation created by the frame rotation into a Fourier-domain modulation.

The Stationary Phase Approximation (SPA) (see e.g.~\cite{Thorne300, CF94}) is a common and very good approximation to compute the Fourier transform of non-precessing signals in the inspiralling phase, but applying it directly to the modulated signal is prone to pathologies. In Ref.~\cite{KCY13,KCY14}, a formalism (called shifted uniform asymptotics or SUA) was introduced to go beyond the SPA and compute more accurately the modulation in the Fourier domain; however, this formalism still relies on the SPA for the underlying signal, and is as such limited to inspiralling signals. Another intuitive approach consists in using a time-to-frequency correspondence, such as the one given by the SPA, to directly carry over the modulation to the Fourier domain. This corresponds, in a sense to be detailed below, to the zeroth-order approximation of the SUA, and it is used implicitly\footnote{With the difference that directly frequency-dependent expressions for the modulation are used, which are then extended to also cover high frequencies in the merger part of the signal, as we will discuss below.} for the simplified precession of the phenomenological waveforms of Ref.~\cite{Hannam+13}. Consistently extending these previous approaches to the merger-ringdown part of the signal is one of the objectives of this paper.

Prospective parameter recovery for LISA observations of comparable-mass binary black hole coalescences is another frontier of gravitational wave signal modelling, becoming a very timely topic with the upcoming effort for the ESA L3 mission characterization in the next few years. In the past, many LISA data analysis studies used mainly post-Newtonian, inspiral-only models, and were limited to the Fisher-matrix formalism instead of full Bayesian analysis (among numerous works, see e.g. \cite{Cutler97,LISAPE09}). However, the Fisher matrix analysis, although considerably cheaper to perform, has known shortcomings including at high SNR~\cite{Vallisneri08,Cokelaer08,Rodriguez+13}. In Ref.~\cite{Klein+15}, an improved model was used, that included the effects of both spins including the corrections beyond the SPA investigated in~\cite{KCY14}; the limitation to inspiral-only waveforms was compensated by SNR reweighting using full IMR waveforms.

The response of the LISA instrument, by contrast with the response of ground-based detectors, has the specificity of being time-dependent. For a signal accumulating over several months, the motion and change of orientation of the detector constellation along its orbit lead to time variability in the response of the detector, which is crucial for the localization of the gravitational-wave source in the sky. More precisely, these effects can be separated as a modulation coming from the rotation of the constellation, and a delay coming from the motion of the spacecrafts in the wavefront of the incoming wave.

@@

\subsection{Notations and conventions}
\label{subsec:notations}

@@

\subsection{Stationary phase approximation}
\label{subsec:SPA}

We recall here an approximation widely used to model gravitational waves signals emitted by inspiralling binaries, the stationary phase approximation (thereafter SPA). It applies in general to chirping signals, and we refer the reader to~\ref{} for details. One first writes the time-domain signal in amplitude and phase form as $h(t) = a(t) e^{-2i\varphi(t)}$, where for gravitational wave signals $\varphi$ will correspond to the orbital phase, with the orbital frequency being $\omega = \dot{\varphi}$. In order to keep close to the notations used in the gravitational wave literature, we introduced a factor of 2 in the phase of the wave, which is appropriate for the dominant 22 harmonic of the signal. The approximation then applies to signals verifying the conditions
\be\label{eq:conditionSPA}
	\left| \frac{\dot{a}/a}{\omega} \right| \ll 1\,, \quad \left|\frac{\dot{\omega}}{\omega^{2}} \right| \ll 1\,, \quad \left| \frac{(\dot{a}/a)^{2}}{\dot{\omega}} \right| \ll 1 \,.
\ee
Since the integral~\eqref{eq:defFT} defining the Fourier transform is rapidly oscillatory unless the term $2\pi f t$ cancels the evolution of $-2\varphi(t)$, its support is well centered around the point of stationary phase. This determines the time-to-frequency correspondence in the stationary phase approximation, and leads to the definition of this time as an implicit function of frequency by the relation
\be\label{eq:deftfSPA}
	\omega(t_{f}^{\rm SPA}) = \pi  f \,.
\ee
For a chirping signal of increasing phase, $\omega>0$ and $\dot{\omega}>0$, there is a unique point of stationary phase, located in the positive frequency range $f>0$. Using the conditions above, one can formally expand the signal around $t_{f}^{\rm SPA}$ to quadratic order in time, according to
\be
	\tilde{h}_{\rm SPA} (f) \simeq a(t_{f}^{\rm SPA}) \exp\left[2i\pi f t_{f}^{\rm SPA}-2i\varphi(t_{f}^{\rm SPA}) \right]  \int \ud t \, e^{-i \dot{\omega} (t_{f}^{\rm SPA}) (t-t_{f}^{\rm SPA})^{2}} \,.
\ee
Note that to be able to treat the amplitude as a constant in the integral above, we used the third condition in~\eqref{eq:conditionSPA}. The resulting complex Gaussian integral yields\footnote{Note that the expressions given here are valid for the $22$ mode of the waveform.}
\begin{subequations}
\begin{align}
	\tilde{h}_{\rm SPA}(f) &= A_{\rm SPA}(f) e^{-i\Psi_{\rm SPA}(f)} \,, \\
	A_{\rm SPA}(f) &= a(t_{f}^{\rm SPA}) \sqrt{\frac{\pi}{\dot{\omega}(t_{f}^{\rm SPA})}} \,, \\
	\Psi_{\rm SPA}(f) &= 2\varphi(t_{f}^{\rm SPA}) - 2\pi f t_{f}^{\rm SPA} + \frac{\pi}{4} \,.
\end{align}
\end{subequations}
Ref.~\cite{Droz+99} evaluated the first correction to this approximation, within the context of post-Newtonian signals, and found that it can be considered as a term of the fifth post-Newtonian order, beyond the accuracy level of our current best models~\cite{BlanchetLiving}.

To understand the separation of timescales in the problem, it will be useful to have at hand the leading-order scaling laws for an inspiral (at the Newtonian order in the PN language). Although inaccurate for the purpose of waveform modelling, these Newtonian estimates will give a good qualitative description of the order of magnitude of the relevant timescales in the inspiral. For a binary with masses $m_{1}, m_{2}$, we define the total mass $m=m_{1}+m_{2}$ and the symmetric mass ratio $\nu = m_{1}m_{2}/m^{2}$. Introducing a time of coalescence $t_{c}$, the relations between the orbital frequency and phase and the time to coalescence $t_{c} - t$ are then given by
\begin{subequations}
\begin{align}
	\omega(t) &= \left[ \frac{256\nu}{5c^{5}} (Gm)^{5/3} (t_{c}-t) \right]^{-3/8} \,, \\
	\varphi(t) &= -\left[ \frac{c^{3}}{G m \nu^{3/5}} (t_{c}-t) \right]^{5/8} \,.
\end{align}
\end{subequations}
For this Newtonian inspiral, applying the SPA gives for the $22$ mode $\tilde{h}(f) = A_{\rm N}(f)e^{-i\Psi_{\rm N}(f)}$ with
\begin{subequations}
\begin{align}
	A_{N}(f) &= \frac{G^{2}m^{2} \pi}{Dc^{5}} \sqrt{\frac{2\nu}{3}} v^{-7/2}\,,\\
	\Psi_{\rm SPA}^{\rm N}(f) &= 2\pi f t_{0} - \phi_{0} + \frac{3}{128\nu v^{5}} \,, 
\end{align}
\end{subequations}
where $v=(G m \pi f/c^{3})^{1/3}$, $D$ is the luminosity distance to the observer, and where $t_{0}, \phi_{0}$ are constants. It is also customary to rewrite the above relations in terms of the chirp mass $\calM \equiv m\nu^{3/5}$, which is the only mass combination arising in the phase at the leading PN order.

@@

Defining
\be
	h_{d}(t) = h(t+d(t)) \,, \quad s(t) = F(t)h_{d}(t) \,,
\ee
our purpose is to derive the Fourier transform $\tilde{s}(f)$, expressed by means of a Fourier-domain transfer function $\calT$ such that
\be\label{eq:deftransfer}
	\tilde{s}(f) \equiv \calT(f) \tilde{h}(f) \,. 
\ee 
We can first write
\be
	h_{d}(t) = \int \ud f \, e^{-2i\pi f (t+d(t))}\tilde{h}(f) \,,
\ee
and
\begin{align}
	\tilde{s}(f) &= \int \ud t \, e^{2i\pi f t} F(t)  \int \ud f' \, e^{-2i\pi f' (t+d(t))}\tilde{h}(f') \nn\\
	&= \int \ud f' \, \tilde{h}(f-f') \int \ud t \, e^{2i\pi f' t} e^{-2i\pi (f-f') d(t)} F(t) \,.
\end{align}
The last equation can then be rewritten as a kind of convolution with a frequency-dependent Kernel, according to
\be\label{eq:FDkernel}
	\tilde{s}(f) = \int \ud f' \, \tilde{h}(f-f') \tilde{G}(f-f',f') \,,
\ee
where we introduced the frequency-dependent function of time $G(f,t)$, and its Fourier transform in the auxiliary frequency $f'$, denoted by $\tilde{g}(f,f')$, as
\begin{subequations}\label{eq:defg}
\begin{align}
	G(f,t) &= e^{-2i\pi f d(t)} F(t) \,, \\
	\tilde{G}(f,f') &= \int \ud t \, e^{2i\pi f' t} e^{-2i\pi f d(t)} F(t) \,.
\end{align}
\end{subequations}
In the absence of delays $d(t)$, as in the case of precessing binaries, the function $G$ loses its frequency-dependence and becomes a modulation in the form of a function of time $F(t)$, and the result~\eqref{eq:FDkernel} above reduces to the usual convolution theorem for the Fourier transform.

In our approach, two assumptions will be crucial for our approximations to be valid:
\begin{itemize}
	\item the time scales entering the signal $h(t)$ are well separated from the ones entering the modulation and the delay, so that $\tilde{G}(f-f',f')$ has support limited to $[-f_{\rm max},f_{\rm max}]$, with $f_{\rm max}$ small compared the characteristic scale for the variation of the Fourier-domain signal;
	\item the original signal has a smooth amplitude and phase in the Fourier-domain, slowly varying over the scale $f_{\rm max}$, so that a local treatment in $f'$ of the convolution in~\eqref{eq:FDkernel} will be possible.
\end{itemize}
Although well verified for the case of LISA-type detectors, the first assumption can be a limitation of the approximation in the case of preccesing binaries, and we will carefully explore its validity in Sec.~\ref{}.

@@

Plugging this relation in~\eqref{eq:FDkernel}, we obtain
\begin{align}
	\tilde{s}(f) &= \tilde{h}(f) \int \ud f' \, \exp\left[ i f' \frac{\ud \Psi}{\ud f} \right] \tilde{G}(f,f') \nn\\
	&= \tilde{h}(f) G\left( f, -\frac{1}{2\pi} \frac{\ud \Psi}{\ud f} \right) \,,
\end{align}
which motivates the introduction of an effective time, seen as a function of frequency,
\be\label{eq:deftf}
	\tf \equiv -\frac{1}{2\pi} \frac{\ud \Psi}{\ud f} \,,
\ee
This definition is a generalization of the time-to-frequency correspondence at the heart of the SPA, as will be discussed in Sec.~\ref{subsec:linkSPA}, but it is worth noting that here it is defined entirely from the Fourier-domain waveform, and independently of the SPA being valid or not for the underlying signal $\tilde{h}$. As such, this time-of-frequency function also extends to the merger-ringdown part of the signals. Fig.~\ref{fig:tf} shows the time function $\tf$ for the example of an equal-mass waveform. In particular, one should note that $\tf$ is not monotonously increasing with frequency anymore when reaching in the high-frequency part of the waveform, corresponding to the ringdown. This would forbid an unambiguous definition of an inverse fequency-of-time correspondence $f(t)$.

@@

Using this notation, we have simply
\be\label{eq:local}
	\calT_{\rm local}(f) = G(f, \tf) \,.
\ee
The interpretation of this approximation is straightforward: the signal is simply multiplied by the response function evaluated at the time $\tf$. Additionally, when delays are present, the frequency-dependence of the response is evaluated directly at $f$. For precessing binary, this corresponds to the approximation used for the PhenomP approximant~\cite{Hannam+13}, as will be discussed in Sec.~\ref{}.

It is worth noting that a shift in time of the time-domain signal will, by construction, be appropriately propagated to~\eqref{eq:local}. If $h'(t) = h(t -  \Delta t)$ [but the modulation and delays are left unchanged (which is relevant for space-based detectors, not for precessing binaries)], $\Psi'(f) = \Psi(f) - 2\pi f \Delta t$, so that
\be\label{eq:localshifted}
	\tilde{s}'_{\rm local}(f) = \tilde{h}'(f) G(f, \tf + \Delta t) \,. 
\ee
This observation also justifies that the first derivative of the Fourier-domain phase cannot, in general, be considered a small quantity, as it can take arbitrary values for arbitrary shifts in time of the signal. The leading order approximation corresponds to neglecting the curvature in $\Psi(f)$, and yields a result~\eqref{eq:local} where the passage to the Fourier domain is done by simply using a time-to-frequency correspondence~\eqref{eq:deftf}, which motivates the name ``local approximation''.

@@

We will formally Taylor-expand $\tilde{h}(f-f')$ in $f'$ as follows: we write
\begin{subequations}
\begin{align}
	\Psi(f-f') &= \Psi(f) + 2\pi f' \tf + \sum\limits_{p\geq 2}^{N} \frac{(-1)^{p}}{p!} {f'}^{p} \frac{\ud^{p} \Psi}{\ud f^{p}} \,, \label{eq:expandPsi}\\
	A(f-f') &= A(f) \sum\limits_{q\geq 0} \frac{(-1)^{q}}{q!} {f'}^{q} \frac{1}{A}\frac{\ud^{q} A}{\ud f^{q}} \,, \label{eq:expandA}\\
	\tilde{G}(f-f', f') &= \sum\limits_{r\geq 0} \frac{(-1)^{r}}{r!} {f'}^{r} \frac{\partial^{r} }{\partial f^{r}}  \tilde{G}(f,f') \label{eq:expandg}
\end{align}
\end{subequations}
This expansion will only be usable in practice when we will be able to keep only a few terms in each of these expansions. We therefore limited the expansion in $\Psi$ to a finite order $N$, to keep the formula readable. Notice that we expand $\tilde{G}(f-f',f')$ in $f'$ only in its first argument. We can also interchange the $f$-derivatives of $g$ with its Fourier transform.
